---
title: "Covid19 Edition"
output: html_notebook
---

# Task 01, Task 02, Task 03, Task 04

- The first code-chunk in a R Markdown document is called "setup".  I
  usually place this before the content of my notebook starts. In it, I
  load packages, data, set options, etc. which I will need throughout
  the document. 
- In this case, I've included some commentary to help you understand
  what is going on.
- Remember, 80% of time spent on most data analysis projects is spent dealing
  with the data. Also, everyone has agreed to make up that 80% number.
    - We spend soooo much time downloading, munging, joinng, etc.
- This code chunk contains:
    - Task 01
    - Task 02
    - Task 03
    - Task 04

```{r setup}
## INIT ========================================================================
library(knitr)
library(lubridate) ## Yes, this is a real package, and yes, you will need to install it.
library(maps)
library(plotly) ## You are going to need to install this too. Sooo much installation.
library(tidyverse)

## County Grain Data --------------------------------------
## In class we worked with the state-grain data. Today, we
## will focus on the counties in a specific state.

## DATA ========================================================================
## Where to get our Covid 19 Data
data_url_counties <- "https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv"

## Defines the state we are going to focus on.
## PLEASE NOTICE HOW THIS IS CAPITALIZED!
which_state <- "New York"
which_county <- "Albany"



## Task 01: Filter and create region
##
## This dplyr statement should: 1. Import the data defined in
## data_url_counties.  2. Filter on which_state so we only keep one
## state's data.  3. Create a new column, region, which should be a
## lower case version of the state column. We did this in class. Feel
## free to look at that code.  4. Create a new column, subregion,
## which should be a lower case verision of the county column.
##
## Pro Tip: Be sure to you mutate to create new columns.

## YOUR CODE HERE!
CountyData <-
    read_csv(data_url_counties) %>%
    filter(state == which_state) %>%
    mutate(region = tolower(state),
           subregion = tolower(county))



## Task 02: Run this and think about what it is doing.
##
## I think I may have mentionned that most of an analysis project is
## data management. Well, here's a whole bunch of data management. You
## owe me a slice of pizza. I have left comments to help you
## understand what/why I am doin. Please take a few minutes to look this over.
## If you have questions, I'm happy to answer them.
CountyData <-
    CountyData %>%
    ## Removes stump rows. Weirdly, New York County was not given a row. Dunno why.
    filter(county != "Bronx", county != "Kings", county != "Richmond", county != "Queens")  %>%
    ## This is a little cruel. It exagerates the severity of the situation in Queens.
    ## It also creates a new date column called week_of.
    mutate(county = case_when(county == "New York City"~"Queens",TRUE~county),
           fips   = case_when(county == "Queens"~"36081",TRUE~fips),
           week_of = ceiling_date(date, unit="weeks")) %>%
    ## This puts the columns into a nice order.
    select(state,region,county,subregion,fips,week_of,cases,deaths) %>%
    ## These next two steps remove date, and replace it with week_of.
    ## This helps to reduce the noise in the data because of accidents in how
    ## or when data was reported to the state.
    group_by(state,region,county,subregion,fips,week_of) %>%
    summarize(cases = sum(cases),
              deaths = sum(deaths)) %>%
    arrange(state,region,county,subregion,fips,week_of) %>%
    group_by(state,region,county,subregion,fips) %>%
    ## This creates a set of "new cases"/"new deaths" columns.
    ## We will explore this more when we discuss exponential growth.
    ## And, it creates columns for each numeric column that is the
    ## log of that column. Whew. That's a lot to absorb.
    mutate(log_cases  = log(cases),
           log_deaths = log(deaths),
           new_cases  = cases - lag(cases, default = first(cases)),
           new_deaths = deaths - lag(deaths, default = first(deaths)),
           log_new_cases = log(new_cases),
           log_new_deaths = log(new_deaths)) %>%
    ## This just prevents us from struggling with it later.
    ungroup() %>%
    ## This is dumb. This should be a numeric. So, there. Now it is.
    mutate(fips = as.numeric(fips)) %>%
    ## Have you noticed I am very particular about how my data is arranged?
    select(state,
           region,
           county,
           subregion,
           fips,
           week_of,
           cases,log_cases,new_cases,log_new_cases,
           deaths,log_deaths,new_deaths,log_new_deaths) %>%
    ## We cannot have incomplete weeks in our daa, because it could falsely
    ## look like a drop in the exponential pattern. This removes incomplete
    ## weeks.
    filter(week_of <= today())
    

## PLEASE note that we have replaced the "date" column with "week_of" which is the last day of the week (Saturday).
## Look at the value produced by the log of 0. Why is this?
## Aren't you glad I didn't make you write that?



## Task 03: Max Dates
##
## Create a dataframe called MaxWeek.
## In it, calculate the max date, per county, found in this data set.
## This is similar to what we did in class. You will need this
## later. Group by fips. Summarize on the max date. Call this column
## MaxDate
##
## Pro Tip: You are going to need to use the assignment operator (<-)
## to complete this task. This will let you create a new dataframe.

## YOUR CODE HERE!
MaxWeek <-
    CountyData %>%
    group_by(fips) %>%
    summarize(MaxDate = max(week_of))

## Create a seconda dataframe called MaxCountyDeaths.
## In it, calculate the max number of cases in each county. This is
## similar to what you just did. Group by fips. Summarize the maximum
## number of cases. Call this column MaxDeaths.
##
## Pro Tip: You are going to need to use the assignment operator (<-)
## to complete this task. This will let you create a new dataframe.

## YOUR CODE HERE!
MaxCountyDeaths <-
    CountyData %>%
    group_by(fips) %>%
    summarize(MaxDeaths = max(deaths))

## Create a datafrane called TooManyDeaths
## Create one more dataframe, based on CountyData which includes only
## counties which have had achieved 100 or more deaths. We will use
## this as a filter below. Other counties are important, but this
## makes the graphs easier to read.
##
## Pro Tip: You are going to need to use the assignment operator (<-)
## to complete this task. This will let you create a new dataframe.

## YOUR CODE HERE!
TooManyDeaths <-
    MaxCountyDeaths %>%
    filter(MaxDeaths >= 100)


## Task 04: Get Population Data
##
## Below, is some code that will import estimates of New York State
## county populations.  Use an inner_join() to join Population and
## CountyData. I recommend joining on fips.
##
##Pro Tip: Any time you have a standard identification, like a FIPS
## code, use it for your joins. You are less likely to spend an hour
## fighting naming convention differences, capitalization differences,
## etc. Trust me on this one.
##
## Pro Tip: You are going to need to use the assignment operator (<-)
## to complete this task. This will let you create a new dataframe.

## This is what I'm giving you. It will create Population.
Population <-
    read_csv("https://data.ny.gov/api/views/krt9-ym2k/rows.csv?accessType=DOWNLOAD&sorting=true") %>%
    filter(Year == 2019) %>%
    select(fips = "FIPS Code", population = Population)

## YOUR CODE HERE!
CountyData <-
    CountyData %>%
    inner_join(Population, by="fips")



## When you are done with these tasks, re-run the entire code
## code-chunk to make sure it works. To do so, hit the green
## button. The code below will return an error if anything is missing
## or deemed incorrect.

if(!exists("MaxWeek") &
   !exists("maxweek") &
   !exists("max_week") &
   !exists("Max_Week")) stop("I don't see MaxWeek")

if(!exists("CountyData")) stop("I don't see CountyData.")
if(!exists("MaxCountyDeaths")) stop("I don't see MaxCountyDeaths.")
if(!exists("TooManyDeaths")) stop("I don't see TooManyDeaths.")

if(!"region" %in% names(CountyData)) stop("I don't see a region column in CountyData.")
if(!"subregion" %in% names(CountyData)) stop("I don't see a subregion column in CountyData.")
if(!"population" %in% names(CountyData)) stop("I don't see a population column in CountyData.")

## Yeah - It took us nearly 200 lines of code just to get this thing started. Whee data.
## Let's go do something interesting.
```

# About This Lab/Notebook

I hope this lab will be useful to you. I hope it helps you
better understand the news you see every day.

Questions/Answers:

- **Will you receive a grade for this?** Yes, of course. But, if you need help,
  reach out to me.
- **What else should I do with this?** I hope the labs we have completed to
  date may be of help to you at some hypothetical point in the future. In
  contrast, I hope this lab is useful to you in the coming days and
  weeks as we face the challenge presented  by the coronavirus together.

## Lab Goals

- You will work on a real, changing data set.
- You will use the county-grain Covid19 data to assess the status of
  the outbreak here in New York State.
- Honestly, this should connect what we have been doing/learning in
  class and show you how these are genuinely useful skills.
- If this lab doesn't convince you these skills are useful, I give up. :-)

## Review County Data:

It is important to understand/inspect your data.

- **Data Source:** [NYT Covid19 Data](https://github.com/nytimes/covid-19-data/blob/master/README.md)
- **WARNING:** This analysis assigns all cases/deaths in New
  York City to Queens. The raw data source assigns all coronavirus cases in 
  the five boroughs to "New York City", which we cannot graph geographically.
  And, because Queens has more cases than any other borough, this analysis
  attributes all cases/deaths to that borough/county.
   - We did this above in the setup chunk.
- I suggest you review this data and seek to understand what is here,
  the grain of the data, and look for any mistakes, etc.

```{r echo=FALSE}
CountyData
```

In class, we looked at cases. In tonight's lab, we will focus on
deaths. Not because I am morbid (although I am) but because deaths
are, sadly, a better measure of progress than total cases. Our testing
efforts are so woefully incomplete that changes up/down in a specific
county could say more about the availability of tests than anything
else.

# Plotting of Exponential Data!

## Task 05: Plot "deaths" by "week_of"

- This is similar to what we did in class.
- X Axis: week_of, Y Axis: deaths, color = county
- Include only those counties in TooManyDeaths. To do this, use the
  dataframe TooManyDeaths as a filter via an inner_join. You can
  either create a temporary dataframe OR use the piping operators to
  do this all in one step. Your choice.
- Pro Tip: Have you thought about joining on fips?

```{r} 
## YOUR CODE HERE!
CountyData %>%
    inner_join(TooManyDeaths, by="fips") %>%
    ggplot(aes(week_of, deaths, color = county)) +
    geom_point() +
    geom_line()
```

Also, while the numbers for Queens are exaggerated because the include
deaths for the other four burroughs as well, take a moment to
appreciate the staggering scale of this epidemic. Labs rarely call for
a moment of silence. This is an exception.



## Task 06: What is the difference?

Look at my code below. Here, I am filtering where deaths >= 100,
rather than using an inner join. I am then producing the plot I
expected you to write for the previous task. Explain to me how these
two plots differ and why I prefer the plot you (hopefully) wrote over
the plot below, which is terrible.

```{r}
## Look at the results here. Compare to what you did above.  How is
## this graph different? Why do I not like this, even thought it is
## easier to understand/read?
CountyData %>%
    filter(deaths >= 100) %>%
    ggplot(aes(week_of, deaths, color = county)) +
    geom_point() +
    geom_line()
```


## Task 07: Which One Is Better

I want us to compare the rate of change across counties. Can you
explain to me why the second plot is better for comparing counties?

### Actual Numbers Plot

- X Axis: deaths, Y Axis: new_deaths, color = county, 
- Include only those counties in Megadeaths.

```{r} 
CountyData %>%
    inner_join(TooManyDeaths, by="fips") %>%
    ggplot(aes(deaths, new_deaths, color = county)) +
    geom_point() +
    geom_line() +
    ggtitle("New Deaths by Total Deaths")

```

### Log Numbers Plot

- X Axis: deaths, Y Axis: new_deaths, color = county
- Transform the X and Y axis to be log10, but show the actual numbers
  for readability/interpretation.
- Include only those counties in Megadeaths.

```{r} 
CountyData %>%
    inner_join(TooManyDeaths, by="fips") %>%
    ggplot(aes(deaths, new_deaths, color = county)) +
    geom_point() +
    geom_line() +
    scale_x_log10() +
    scale_y_log10()
```

Why is the second plot better for comparing different counties?  Think
back to the video.

# Maps!

## Task 08: New York State

Draw a State/County map of New York State. This is similar to what we
did last week in class, but I'm adding the complexity of plotting the
counties as well.

Review this map and tell me what you think it is telling you in Canvas.

```{r}

## First, let's load some map data and look at it.
## You only need to load this in once per sussion.
StateMap <- map_data("state") %>% filter(region == "new york")
CountyMap <- map_data("county") %>% filter(region == "new york")

## Now let's build our data set.
## See how we are re-using our MaxDates data?
## And . . . . see how this data transformation stuff IS useful?
## When doing REAL statistics.

## This steop takes a minute.
CountyMapData <- 
    CountyData %>%
    inner_join(MaxWeek, by = c("fips","week_of"="MaxDate")) %>%
    inner_join(CountyMap, by = "subregion")

ggplot(CountyMapData, aes(long, lat, group = group))+
    geom_polygon(aes(fill = deaths), color = "white")+
    theme_classic() +
    ggtitle("Distribution of Deaths by Coronavirus (2020)")
```

## Task 09: Deaths Per 1000 County Residents

- There is some auto-correlation between population size (of the
  county) and the number of coronavirus cases there are. All else
  being equal, larger counties will tend to have more deaths.
- Add a column to CountyData called prop, for proportion. This should
  be the number of deaths per 10,000 county residents.
- Remember, you have to assign the output of your dplyr mutate to
  CountyData or it won't work.
- Then, draw a new map showing the distribution of this new column,
  like what we did above.

```{r}
## YOUR CODE HERE!
CountyData <-
    CountyData %>%
    mutate(prop = 10000*deaths/population)

## This steop takes a minute.
CountyMapData <- 
    CountyData %>%
    inner_join(MaxWeek, by = c("fips","week_of"="MaxDate")) %>%
    inner_join(CountyMap, by = "subregion")

ggplot(CountyMapData, aes(long, lat, group = group))+
    geom_polygon(aes(fill = prop), color = "white")+
    theme_classic() +
    ggtitle("Distribution of Deaths by Coronavirus (2020)")
```

Hopefully, you can see that although the distribution of raw case
counts is very biased to NYC, the distribution of prop is much wider
and it makes all of look quite a bit different.

Upload this image into Canvas.

# Task 10: Is our work reproducible?

- Assuming the New York Times leaves the data on their website, is this lab reproducible next year?
- Is the code available for inspection?
- Is the analysis connected to the code?
- Do you think this sort of scientific workflow would help us address the reproducibility crisis?
    - I obviously think it can. However, feel free to disagree with me
      if that's how you feel. I want you to think about this, not
      fawningly agree with me to get a good grade.

# Task 11: Upload Your R Script

# Task 12: Was this lab useful?

Was this lab useful? This is an "essay" question, but I want your
honest thoughts/opinions. Any reasonable answer will receive full
credit.

# Task 13: Lucky!

Hit the preview button to be sure your code all works (RStudio will throw an 
error if you have any problems). Once that is done, upload the file 
`County Grain Coronavirus Cases.Rmd`. Do not upload the .Rproj file. That is
something else and is not useful to me from a grading standpoint.

See how by default the code chunks are hidden but can be expanded/shown with 
the click of a button.

# Task 14: Extra Credit

If you can figure out how to apply all of these graphs to NYS counties
with fewer than 100 coronavirus deaths, I'll give you half a
letter-grade on your lowest non-exam grade in the class. If you can
figure out how to run the code against another state, there are even
more opportunities:

- Figure out how to run everything except the population data
  transformations/graph and I'll give you a full letter grade on your
  lowest non-exam grade.
- Figure out how to run the population transformations/graphs for that
  state as well, and I'll just give you an A on your lowest non-exam
  grade in the class.
  
This latter opportunity should be of interest to those of you with an
incomplete item or two.
