---
title: "Covid19 Edition"
output: html_notebook
---

# Task 01, Task 02, Task 03, Task 04

- The first code-chunk in a R Markdown document is called "setup".  I
  usually place this before the content of my notebook starts. In it, I
  load packages, data, set options, etc. which I will need throughout
  the document. 
- In this case, I've included some commentary to help you understand
  what is going on.
- Remember, 80% of time spent on most data analysis projects is spent dealing
  with the data. Also, everyone has agreed to make up that 80% number.
    - We spend soooo much time downloading, munging, joinng, etc.
- This code chunk contains:
    - Task 01
    - Task 02
    - Task 03
    - Task 04

```{r setup}
## INIT ========================================================================
library(knitr)
library(lubridate) ## Yes, this is a real package, and yes, you will need to install it.
library(maps)
library(tidyverse)
options(scipen=999)

## County Grain Data --------------------------------------
## In class we worked with the state-grain data. Today, we
## will focus on the counties in a specific state.

## DATA ========================================================================
## Where to get our Covid 19 Data
data_url_counties <- "https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv"

## Defines the state we are going to focus on.
## PLEASE NOTICE HOW THIS IS CAPITALIZED!
which_state <- "New York"
which_county <- "Albany"



## Task 01: Filter and create region
##
## This dplyr statement should: 
## 1. Import the data defined in data_url_counties.  
## 2. Filter on which_state (keep only NYS data)
## 3. Create a new column, region, which should be a
##    lower case version of the state column. We did this in class.
##    Feel free to look at that code.  4. Create a new column, subregion,
##    which should be a lower case verision of the county column.
##
## Pro Tip: Be sure to you mutate to create new columns.
##
## Copy/Paste your code to Canvas when done.

## YOUR CODE HERE!



## Task 02: Run this and think about what it is doing.
##
## I think I may have mentionned that most of an analysis project is
## data management. Well, here's a whole bunch of data management. You
## owe me a slice of pizza. I have left comments to help you
## understand what/why I am doing. Please take a few minutes to look this over.
## If you have questions, I'm happy to answer them.
CountyData <-
    CountyData %>%
    ## Removes stump rows. Weirdly, New York County was not given a row. Dunno why.
    filter(county != "Bronx", county != "Kings", county != "Richmond", county != "Queens")  %>%
    ## This is a little cruel. It exagerates the severity of the situation in Queens.
    ## It also creates a new date column called week_of.
    mutate(county = case_when(county == "New York City"~"Queens",TRUE~county),
           fips   = case_when(county == "Queens"~"36081",TRUE~fips),
           week_of = ceiling_date(date, unit="weeks")) %>%
    ## This puts the columns into a nice order.
    select(state,region,county,subregion,fips,week_of,cases,deaths) %>%
    ## These next two steps remove date, and replace it with week_of.
    ## This helps to reduce the noise in the data because of accidents in how
    ## or when data was reported to the state.
    group_by(state,region,county,subregion,fips,week_of) %>%
    summarize(cases = last(cases),
              deaths = last(deaths)) %>%
    arrange(state,region,county,subregion,fips,week_of) %>%
    ## This just prevents us from struggling with it later.
    ungroup() %>%
    ## This creates a set of "new cases"/"new deaths" columns.
    ## We will explore this more when we discuss exponential growth.
    ## And fips should have already been a numeric. Sheesh
    mutate(new_cases  = cases - lag(cases, default = first(cases)),
           new_deaths = deaths - lag(deaths, default = first(deaths)),
           fips       = as.numeric(fips)) %>%
    ## Have you noticed I am very particular about how my data is arranged?
    select(state,
           region,
           county,
           subregion,
           fips,
           week_of,
           cases,new_cases,
           deaths,new_deaths) %>%
    ## We cannot have incomplete weeks in our daa, because it could falsely
    ## look like a drop in the exponential pattern. This removes incomplete
    ## weeks.
    filter(week_of <= today())
    
## PLEASE note that we have replaced the "date" column with "week_of" which is the last day of the week (Saturday).
## Look at the value produced by the log of 0. Why is this?
## Aren't you glad I didn't make you write that?



## Task 03: Max Dates
##
## Create a dataframe called MaxDate.
## In it, calculate the max date, per county, found in this data set.
## This is similar to what we did in class. You will need this
## later. Group by fips. Summarize on the max date. Call this column
## max_date
##
## Pro Tip: Group by fips. Not county. FIPS Codes are a way to identify
## counties and they will make our joins later less painful.
##
## Pro Tip: You are going to need to use the assignment operator (<-)
## to complete this task. This will let you create a new dataframe.
##
## In Canvas, tell me what the MaxDate for fips 36001 is.

## YOUR CODE HERE!

## Create a second dataframe called MaxCases.
## In it, calculate the max number of cases in each county. This is
## similar to what you just did. Group by fips. Summarize the maximum
## number of cases. Call this column MaxCountyCases.
##
## Pro Tip: Group by fips. Not county. FIPS Codes are a way to identify
## counties and they will make our joins later less painful.
##
## Pro Tip: You are going to need to use the assignment operator (<-)
## to complete this task. This will let you create a new dataframe.
## 
## In Canvas, tell me the MaxCases value for fips 36001.

## YOUR CODE HERE!

## Create a datafrane called Over1000
## Create one more dataframe, based on MaxCountyCases which includes only
## counties which have had 1000 or more max cases. We will use
## this as a filter below. Other counties are important, but this
## makes the graphs easier to read.
##
## Pro Tip: You are going to need to use the assignment operator (<-)
## to complete this task. This will let you create a new dataframe.
##
## In Canvas, tell me how many counties are in Over1000 (how many rows)?

## YOUR CODE HERE!


## Task 04: Get Population Data
##
## Below, is some code that will import estimates of New York State
## county populations. Use an inner_join() to join Population and
## CountyData. I recommend joining on fips. You want to create a new
## CountyCases dataframe that ALSO contains the population data.
##
## Pro Tip: Any time you have a standard identification, like a FIPS
## code, use it for your joins. You are less likely to spend an hour
## fighting naming convention differences, capitalization differences,
## etc. Trust me on this one.
##
## Pro Tip: You are going to need to use the assignment operator (<-)
## to complete this task. This will let you create a new dataframe.
##
## In Canvas, tell me the estimated 2019 population of Albany County.

## This is what I'm giving you. It will create Population.
Population <-
    read_csv("https://data.ny.gov/api/views/krt9-ym2k/rows.csv?accessType=DOWNLOAD&sorting=true") %>%
    filter(Year == 2019) %>%
    select(fips = "FIPS Code", population = Population)

## YOUR CODE HERE!
## You need to do an inner_join here.


```

# About This Lab/Notebook

I hope this lab will be useful to you. I hope it helps you
better understand the news you see every day.

Questions/Answers:

- **Will you receive a grade for this?** Yes, of course. But, if you need help,
  reach out to me.
- **What else should I do with this?** I hope the labs we have completed to
  date may be of help to you at some hypothetical point in the future. In
  contrast, I hope this lab is useful to you in the coming days and
  weeks as we face the challenge presented  by the coronavirus together.

## Lab Goals

- You will work on a real, changing data set.
- You will use the county-grain Covid19 data to assess the status of
  the outbreak here in New York State.
- Honestly, this should connect what we have been doing/learning in
  class and show you how these are genuinely useful skills.
- If this lab doesn't convince you these skills are useful, I give up. :-)

## Review County Data:

It is important to understand/inspect your data.

- **Data Source:** [NYT Covid19 Data](https://github.com/nytimes/covid-19-data/blob/master/README.md)
- **WARNING:** This analysis assigns all cases/deaths in New
  York City to Queens. The raw data source assigns all coronavirus cases in 
  the five boroughs to "New York City", which we cannot graph geographically.
  And, because Queens has more cases than any other borough, this analysis
  attributes all cases/deaths to that borough/county.
   - We did this above in the setup chunk.
- I suggest you review this data and seek to understand what is here,
  the grain of the data, and look for any mistakes, etc.

```{r echo=FALSE}
CountyData
```

We will continue to look at and focus on the growth and distribution of cases
throughout New York State.

# Plotting of Exponential Data!

## Task 05: Plot "cases" by "week_of"

- This is similar to what we did in class.
- X Axis: week_of, Y Axis: cases, color = county
- Include only those counties in Over1000. To do this, use the
  dataframe Over1000 as a filter via an inner_join. You can
  either create a temporary dataframe OR use the piping operators to
  do this all in one step. Your choice.
- Pro Tip: Have you thought about joining on fips?
- Upload this to Canvas.

```{r} 
## YOUR CODE HERE!

```

Also, while the numbers for Queens are exaggerated because the include
cases for the other four burroughs as well, take a moment to
appreciate the staggering scale of this epidemic. Labs rarely call for
a moment of silence. This is an exception.



## Task 06: What is the difference?

Look at my code below. Here, I am filtering where cases >= 1000,
rather than using an inner join. I am then producing the plot I
expected you to write for the previous task. Explain to me how these
two plots differ and why I prefer the plot you (hopefully) wrote over
the plot below, which is terrible.

```{r}
## Look at the results here. Compare to what you did above.  How is
## this graph different? Why do I not like this, even thought it is
## easier to understand/read? In what way(s) is it incorrect or 
## omit useful information?
CountyData %>%
    filter(cases >= 1000) %>%
    ggplot(aes(week_of, cases, color = county)) +
    geom_point() +
    geom_line()
```


## Task 07: Which One Is Better

I want us to compare the rate of change across counties. Below, I have 
provided you with two plots. Can you explain to me why the second plot
is better for comparing counties?

### Actual Numbers Plot

- X Axis: cases, Y Axis: new_cases, color = county, 
- Include only those counties in Over1000.

```{r} 
CountyData %>%
    inner_join(Over1000, by="fips") %>%
    ggplot(aes(cases, new_cases, color = county)) +
    geom_point() +
    geom_line() +
    ggtitle("New Cases by Current Cases")

```

### Log Numbers Plot

- X Axis: cases, Y Axis: new_cases, color = county
- Transform the X and Y axis to be log10, but show the actual numbers
  for readability/interpretation.
- Include only those counties in Over1000.

```{r} 
CountyData %>%
    filter(new_cases > 0) %>%
    inner_join(Over1000, by="fips") %>%
    ggplot(aes(x = cases, y = new_cases, color = county)) +
    geom_point() +
    geom_line() +
    scale_x_log10() +
    scale_y_log10()
```

Why is the second plot better for comparing different counties?  **Hint:**
Think back to the video.

# Maps!

## Task 08: New York State

Draw a State/County map of New York State. This is similar to what we
did last week in class, but I'm adding the complexity of plotting the
counties as well.

Review this map and tell me what you think it is telling you in Canvas.

```{r}

## First, let's load some map data and look at it.
## You only need to load this in once per sussion.
StateMap <- map_data("state") %>% filter(region == "new york")
CountyMap <- map_data("county") %>% filter(region == "new york")

## Now let's build our data set.
## See how we are re-using our MaxDates data?
## And . . . . see how this data transformation stuff IS useful?
## When doing REAL statistics.

## This step takes a minute.
CountyMapData <- 
    CountyData %>%
    inner_join(MaxDate, by = c("fips","week_of"="max_date")) %>%
    inner_join(CountyMap, by = "subregion")

ggplot(CountyMapData, aes(long, lat, group = group))+
    geom_polygon(aes(fill = cases), color = "white")+
    theme_classic() +
    ggtitle("Distribution of Coronavirus Cases (2020)")
```

## Task 09: Cases Per 1000 County Residents

- There is some auto-correlation between population size (of the
  county) and the number of coronavirus cases there are. All else
  being equal, larger counties will tend to have more cases.
- Add a column to CountyData called prop, for proportion. This should
  be the number of cases per 1000 county residents.
- Remember, you have to assign the output of your dplyr mutate to
  CountyData or it won't work.
- Then, draw a new map showing the distribution of this new column,
  like what we did above.
- Upload this map to Canvas.

```{r}
## YOUR CODE HERE!

```

Hopefully, you can see that although the distribution of raw case
counts is very biased to NYC, the distribution of prop is much wider
and it makes all of look quite a bit different.

Now do the same for deaths. Do you see a similar pattern? Are deaths per 1,000
residents similarly distributed?

# Task 10: Is our work reproducible?

- Assuming the New York Times leaves the data on their website, is this lab reproducible next year?
- Is the code available for inspection?
- Is the analysis connected to the code?
- Do you think this sort of scientific workflow would help us address the reproducibility crisis?
    - I obviously think it can. However, feel free to disagree with me
      if that's how you feel. I want you to think about this, not
      blindly agree with me to get a good grade.

# Task 11: Upload Your R Script

Yeah. Just do what the title says. Upload the `Lab 11 - Codid19 Edition.Rmd` file.

# Task 12: Was this lab useful?

Was this lab useful? This is an "essay" question, but I want your
honest thoughts/opinions. Any reasonable answer will receive full
credit.

# Task 13: Extra Credit

If you can figure out how to apply all of these graphs to NYS counties
with fewer than 1,000 coronavirus cases, I'll give you half a
letter-grade on your lowest non-exam grade in the class. If you can
figure out how to run the code against another state, there are even
more opportunities:

- Figure out how to run everything except the population data
  transformations/graph and I'll give you a full letter grade on your
  lowest non-exam grade.
- Figure out how to run the population transformations/graphs for that
  state as well, and I'll just give you an A on your lowest non-exam
  grade in the class.
  
This latter opportunity should be of interest to those of you with an
incomplete item or two.


```{r}
CountyData %>%
    filter(new_cases > 0, county %in% c("Albany","Saratoga","Rensselaer","Schenectady","Greene","Columbia","Schoharie","Montgomery","Washington","Warren","Fulton")) %>%
    mutate(county = "Capital District") %>%
    ggplot(aes(x = cases, y = new_cases, color = county)) +
    geom_point() +
    geom_line() +
    scale_x_log10() +
    scale_y_log10()
```

