---
title: "Insurance T-Tests"
output: html_notebook
---

# Insurance

Do you remember the Insurance data set we played with back before the 
Coronapocalypse? Well, it's back! I'm going to give you a break this week. 
This should be a fairly easy lab.

Because you will need to download (or re-download) the Insurance data, I suggest
using R Studio to create a new project for this lab or re-use the project you
created for lab 3. Your choice. Either way, you will need to have the file,
`insurance.csv` in the same folder as this R Markdown Notebook.

# Task 01: Import data

- Run this first code chunk. It imports the data. If it cannot find the data, 
  it will return an error.
- In Canvas, tell me how many rows of data are in `insurance.csv`.

```{r}
library(tidyverse)
options(scipen = 999)


## DATA =======================================================================
## Loads the data and provides an initial view.  This code checks for
## the presence of insurance.csv. If it fails to find the data file,
## it will give you an error and stop.
data_file <- "insurance.csv"
if (file.exists(data_file)){
    Insurance <- read_csv(file = data_file)
} else {
    stop("The data file is not in your current working directory.")
}
Insurance

```



# Task 02: BMI by Gender

- Add a new code-chunk by clicking the *Insert Chunk* button on the toolbar or 
  by pressing *Ctrl+Alt+I*.
- Calculate:
    - the average BMI
    - the standard deviation of BMI
    - stratify by gender `sex`
- In Canvas, tell me which gender has the highest average BMI.
- *Pro Tip:* Leave a blank line between any/all plain text and the fence 
  markers which surround a code-chunk. 

YOUR CODE CHUNK HERE!

```{r}
Insurance %>%
    group_by(sex) %>%
    summarize(n = n(), AvgBMI = mean(bmi), SDBMI = sd(bmi))
```



# Task 03: T-Test, Gender

- The t-test can only compare two averages at a time.
- Are the observed differences in BMI (less than 1) between the two genders
  statistically significant? Use the `t.test` to calculate the p-value and 
  reject the null hypothesis if p <= .05.
- In Canvas: 
    - Tell me what the null hypothesis is in this example. (No more than a few 
      sentences.)
    - Tell me the p-value. (you can round to two decimal places.)
    - Tell me if you can can reject the null hypothesis.
- Hint: We discussed, at length, how to use R's formula notation.
    - In this case, BMI as a function of Gender.
    - Now, how do you write that in the t.test parameters?
- Pro Tip: If you use `%>%` to pass the data to the `t.test` function, you must
  manually include a parameter `data = .`. Because t.test is an OLD SKOOL R 
  function, it cannot magically find the data. We have to be explicit.
  
YOUR CODE CHUNK HERE!

```{r}
Insurance %>% t.test(bmi~sex, data=.)
```



# Task 04: Do You Agree?

- Insert a code chunk.
- Draw a box-plot of BMI by sex. Upload this to Canvas.
- Do you agree/disagree with the results of the t-test above? 
  Tell me why/why not.




# Task 05: How many rows would we need?

- See the documentation for `power.t.test` for more information.
    - Hint: `?power.t.test` in the console.
    - OR use the help section in the pane to the right -->
- This function has five parameters which interact. When used, exactly
  one parameter must be passed as NULL.
- I would like to estimate how much data I need to collect in order to find
  statistically significant results.
- Use the following values in `power.t.test`.
    - n: NULL
    - delta: The observed difference between the BMI of men and women.
    - power: .9 (This is standard, for Type II error)
    - sd = The observed standard deviation for the two groups. They are similar.
    - sig.level: .05 (This is a standard, for Type I error)
- Use your calculations from Task 02 to help you complete this.
- When calculating values for delta and sd, don't go beyond 2 decimal places.
- Also, feel free to use a calculator. You don't have to do this in R.
- For sd, use the bigger of the two to be conservative.
- To better understand power and sig.level, read this: [Type I and Type II Errors](https://en.wikipedia.org/wiki/Type_I_and_type_II_errors)

YOUR CODE CHUNK HERE!


NOTE: THE ANSWER HERE IN CANVAS IS WRONG!!!!!

```{r}
power.t.test(n = NULL, 
             delta = .57, 
             power = .9, 
             sd = 6.14, 
             sig.level = .05)
```



# Task 06: BMI by Region

- Calculate:
    - the average BMI
    - the standard deviation of BMI
    - stratify by region
- In Canvas, tell me which region has the highest average BMI.

YOUR CODE CHUNK HERE!

```{r}
Insurance %>%
    group_by(region) %>%
    summarize(n = n(), AvgBMI = mean(bmi), SDBMI = sd(bmi))
```



# Task 07: T-Test, Region

- Is the observed difference between the BMI in the South East ans South West 
  statistically significant?
- You will need to create a code-chunk, and filter the data down to only those 
  two regions.
    - This is something we haven't done much of.
    - In you filter command, do something like `region %in% c("southeast","southwest")`
- In Canvas: 
    - Tell me what the null hypothesis is in this example. (No more than a few 
      sentences.)
    - Tell me the p-value. (you can round of to two decimal places.)
    - Tell me if you can can reject the null hypothesis.
- Pro Tip: If you use `%>%` to pass the data to the `t.test` function, you must
  manually include a parameter `data = .`. Because t.test is an OLD SKOOL R 
  function, it cannot magically find the data. We have to be explicit.

YOUR CODE CHUNK HERE!

```{r}
Insurance %>% 
  filter(region %in% c("southeast","southwest")) %>% 
  t.test(bmi~region, data=.)
```



# Task 08: Minimum number of rows?

- See the documentation for `power.t.test` for more information.
    - Hint: `?power.t.test` in the console.
    - OR use the help section in the pane to the right -->
- This function has five parameters which interact. When used, exactly
  one parameter must be passed as NULL.
- I would like to estimate the minimum number of rows needed in order to find
  statistically significant results between these two regions.
- Use the following values in `power.t.test`.
    - n: NULL
    - delta: The observed difference between the BMI of the southeast and southwest.
    - power: .9 (This is standard, for Type II error)
    - sd = The observed standard deviation for the two groups. They are similar.
    - sig.level: .05 (This is a standard, for Type I error)
- Use your calculations from Task 02 to help you complete this.
- When calculating values for delta and sd, don't go beyond 2 decimal places.
- Also, feel free to use a calculator. You don't have to do this in R.
- For sd, use the bigger of the two to be conservative.

YOUR CODE CHUNK HERE!

```{r}
power.t.test(n = NULL, 
             delta = 3.2, 
             power = .9, 
             sd = 6.5, 
             sig.level = .05)
```



# Task 09: Next Steps

- I _hope_ I have proven that achieving statistical significance is about
  the interaction between the observed difference (delta), the standard 
  deviation, amount of data collected, and power/significance level.
    - A smaller standard deviation will reduce the amount of data needed to
      achieve statistical significance.
    - A larger difference in observed means will decrease the amount of data
      needed to achieve statistical significance.
- One major concern I have is how statistical significance is often applied.
    - Researchers rarely determine how much data they will collect. The use
      a convenience sample from an EHR, or they estimate how many people they
      can test based on the value of their latest grant.
    - And, in all fairness, more data is always better if the data is unbiased.
- Given all of these constraints, what can researchers do to protect
  themselves (and the public) from reporting false positives?
- Hint: This isn't something that can be solved for in R. This is a thought 
  experiment.
  


# Task 10: Upload your .Rmd file to Canvas

- You know the drill.
- BE CAREFUL to not upload Insurance.R if you reused the Lab 3 project folder.