---
title: "Lucky Lab"
output: html_notebook
---

# What Is Good?

I know these last few weeks have been incredibly tough.

[The Story of the Old Farmer](http://emptygatezen.com/teaching/2017/4/7/story-of-the-old-farmer)

# Our Data - Births

We were first introduced to the births data in a lab before Spring Break. (I 
know, that seems like eons ago.) The data here is anonymous, but comes from the
Capital District. It is also real, although old, data.

```{r}
## INIT ========================================================================
library(tidyverse)
options(scipen = 999)

births_url <- "https://github.com/intro-to-data/Data/raw/master/births.csv"
Births <- read_csv(births_url)
names(Births) <- c("ID", "Facility", "Insurance", "Gender", "LOS", "Admitted", "Discharged", "BirthWeight", "TotalCharges")
Births
```

It is always necessary to understand the context of the data. Below, is some
information about our data:

| # | Column      | Column Description                                           |
|--:|:------------|:-------------------------------------------------------------|
| 1 | ID          | Patient ID: Not an actual MRN or MemberID.                   |
| 2 | Facility    | Hospital where the birth took place in the Capital District. |
| 3 | Insurance   | Which insurance company paid for the birth?                  |
| 4 | Gender      | Gender of baby.                                              |
| 5 | LOS         | Length of Stay in days.                                      |
| 6 | Admitted    | Day mother was admitted to hospital.                         |
| 7 | Discharged  | Day baby was discharged from hospital.                       |
| 8 | BirthWeight | Birth Weight (grams).                                        |
| 9 | TotalCharges| Total Charges for the birth.                                 |

This data should be familiar. Like your favorite blanket.



# Tasks



## Task 01: Charges

- Plot a density plot of `TotalCharges`.
- In Canvas, how would you describe the distribution of `TotalCharges`?
  - Normally distributed, skewed?
- In Canvas upload the plot of `TotalCharges` density plot.

```{r}
## YOUR CODE HERE!

Births %>% 
    ggplot(aes(TotalCharges)) +
    geom_density() +
    ggtitle("Density, Total Charges")
```



## Task 02: LOS

- Plot `TotalCharges~LOS`. That is, draw a plot where `LOS` is the independent
  variable and `TotalCharges` is the dependent variable. 
- Include a `geom_smooth(method="lm")` in your plots.
- In Canvas, describe your observations about the relationship between these 
  two variables. Include:
  - Is the relationship positive or negative?
  - Do the residuals appear to be normally distributed?
  - Does the relationship actually appear linear? (And not curved, which would 
    suggest a more complex model.)
  - Later we will try to produce a model predicting total charges. Based on your 
    observations, which will this feature be useful in our model?

```{r}
## YOUR CODE HERE!

Births %>% 
    ggplot(aes(LOS, TotalCharges)) +
    geom_point() +
    geom_smooth(method = "lm") +
    ggtitle("TotalCharges~LOS")
```




## Task 03: Birth Weight

- Plot `TotalCharges~BirthWeight`. That is, draw a plot where `BirthWeight` is 
  the independent variable and `TotalCharges` is the dependent variable.
- Include a `geom_smooth(method="lm")` in your plot.
- In Canvas, describe your observations about the relationship between these 
  two variables. Include:
  - Is the relationship positive or negative?
  - Do the residuals appear to be normally distributed?
  - Does the relationship actually appear linear? (And not curved, which would 
    suggest a more complex model.)
  - Later we will try to produce a model predicting total charges. Do you think
    this feature would be useful?

```{r}
## YOUR CODE HERE!
  
Births %>% 
    ggplot(aes(BirthWeight, TotalCharges)) +
    geom_point() +
    geom_smooth(method = "lm") +
    ggtitle("TotalCharges~BirthWeight")
```



## Task 04: First Linear Model

- Create a linear model of `TotalCharges~LOS`. Call it `TCLS`.
- Pro Tip: The function for creating a linear model is `lm`. The function to
  see the model details, use `summary(TCLS)`.
- In Canvas:
  - Enter the R-squared for `TCLS`.

```{r}
## YOUR CODE HERE (TCLS)
TCLS <- lm(TotalCharges~LOS, data = Births)
summary(TCLS)
```



## Task 05: Second Linear Model

- Create a linear model of `TotalCharges~BirthWeight`. Call it `TCBW`.
- In Canvas:
  - Enter the R-squared for `TCBW`.

```{r}
## YOUR CODE HERE (TCBW)
TCBW <- lm(TotalCharges~BirthWeight, data = Births)
summary(TCBW)
```

## Task 06: Which is a better predictor of Total Charges

- Based on your understanding of how the linear model works, which of the two
  models, `TCLS` or `TCBW`, is a better predictor of TotalCharges?
- Tell me why in Canvas. Integrate information from your plots and your model 
  to support your argument (you do not need to upload the image, unless you 
  just want to). I know what the plots look like, so you can describe them.



## Task 07: Combined

- Create one final model, which uses `LOS` and `BirthWeight`, to predict Total Charges.
- Call this model `Combined`.
- Copy/paste the code you settle on to produce this model in Canvas.

```{r}
## YOUR CODE HERE!


```


- Based on our discussion of R model formulas last week, create a single linear
  model that uses both LOS and 


